---
title: "MethylKit"
author: "Yaamini Venkataraman"
date: "10/25/2018"
output: html_document
---

In this file, I'll identify differentially methylated loci (DML) using `methylKit`.

# Prepare R Markdown file

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Install packages

```{r}
install.packages("devtools") #Install the devtools package
library(devtools) #Load devtools

source("https://bioconductor.org/biocLite.R") #Source package from bioconductor
biocLite("methylKit") #Install methylkit
```

```{r}
install_github("al2na/methylKit", build_vignettes = FALSE, repos = BiocInstaller::biocinstallRepos(), dependencies = TRUE) #Install more methylKit options
library(methylKit) #Load methylkit
```

# Obtain session information

```{r}
sessionInfo()
```

# Download files from `gannet`

```{bash}
wget -r -l1 --no-parent -A_dedup.sorted.bam http://gannet.fish.washington.edu/spartina/2018-10-10-project-virginica-oa-Large-Files/2018-11-07-Bismark-Mox/ #Download files from gannet. They will be stored in the same directory structure as they are online.
```

```{bash}
mv gannet.fish.washington.edu/spartina/2018-10-10-project-virginica-oa-Large-Files/2018-11-07-Bismark-Mox/* analyses/2018-10-25-MethylKit #Move all files from gannet folder to analyses folder
```

```{bash}
rm -r gannet.fish.washington.edu #Remove the empty gannet directory
```

# Process methylation data

```{r}
analysisFiles <- list("zr2096_1_dedup.sorted.bam",
                      "zr2096_2_dedup.sorted.bam",
                      "zr2096_3_dedup.sorted.bam",
                      "zr2096_4_dedup.sorted.bam",
                      "zr2096_5_dedup.sorted.bam",
                      "zr2096_6_dedup.sorted.bam",
                      "zr2096_7_dedup.sorted.bam",
                      "zr2096_8_dedup.sorted.bam",
                      "zr2096_9_dedup.sorted.bam",
                      "zr2096_10_dedup.sorted.bam") #Put all .bam files into a list for analysis.
```

```{r}
sample.IDs <- list("1", "2", "3", "4", "5", "6", "7", "8", "9", "10") #Create list of sample IDs
treatmentSpecification <- c(rep(0, times = 5), rep(1, times = 5)) #Specify which treatment the samples were from. 1 is the treatment (high pCO2) and 0 is the control (ambient pCO2)
```

```{r}
processedFilesCov5 <- processBismarkAln(location = analysisFiles, sample.id = sample.IDs, assembly = "v3", read.context = "CpG", mincov = 5, treatment = treatmentSpecification) #Process files for CpG methylation using 5x coverage. First 5 files were from ambient conditions, and the second from high pCO2 conditions.
```

# Differentially methylated loci

## Obtain methylation and coverage plots

```{r}
nFiles <- length(sample.IDs) #Count number of samples
fileName <- data.frame("nameBase" = rep("2018-10-25-Loci-Analysis/2018-11-07-Percent-CpG-Methylation", times = nFiles),
                       "nameBase2" = rep("2018-10-25-Loci-Analysis/2018-11-07-Percent-CpG-Coverage", times = nFiles),
                       "sample.ID" = 1:10) #Create new dataframe for filenames
head(fileName) #Confirm dataframe creation
```

```{r}
fileName$actualFileName3 <- paste(fileName$nameBase, "-5xCoverage", "-Sample", fileName$sample.ID, ".jpeg", sep = "") #Create a new column for the full filename for 5x coverage + specific sample's percent CpG methylation plot
fileName$actualFileName4 <- paste(fileName$nameBase2, "-5xCoverage", "-Sample", fileName$sample.ID, ".jpeg", sep = "") #Create a new column for the full filename for 5x coverage + specific sample's percent CpG coverage plot
head(fileName) #Confirm column creation
```

```{r}
for(i in 1:nFiles) { #For each data file
  jpeg(filename = fileName$actualFileName3[i], height = 1000, width = 1000) #Save file with designated name
  getMethylationStats(processedFilesCov5[[i]], plot = TRUE, both.strands = FALSE) #Get %CpG methylation information
  dev.off() #Turn off plotting device
} #Plot and save %CpG methylation information
```

```{r}
for(i in 1:nFiles) { #For each data file
  jpeg(filename = fileName$actualFileName4[i], height = 1000, width = 1000) #Save file with designated name
  getCoverageStats(processedFilesCov5[[i]], plot = TRUE, both.strands = FALSE) #Get CpG coverage information
  dev.off() #Turn off plotting device
} #Plot and save CpG coverage information
```

## Obtain clustering information

```{r}
methylationInformationCov5Destrand <- unite(processedFilesCov5, destrand = TRUE) #Combine all processed files into a single table. Include destrand = TRUE to remove all strand information.
head(methylationInformationCov5Destrand) #Confirm unite
```

```{r}
clusteringInformationCov5Destrand <- clusterSamples(methylationInformationCov5Destrand, dist = "correlation", method = "ward", plot = FALSE) #Save cluster information as a new object
```

```{r}
jpeg(filename = "2018-10-25-Loci-Analysis/2019-03-14-Full-Sample-Pearson-Correlation-Plot-Cov5Destrand.jpeg", height = 1000, width = 1000) #Save file with designated name
getCorrelation(methylationInformationCov5Destrand, plot = TRUE) #Understand correlation between methylation patterns in different samples
dev.off()
```

```{r}
jpeg(filename = "2018-10-25-Loci-Analysis/2019-03-14-Full-Sample-CpG-Methylation-Clustering-Cov5Destrand.jpeg", height = 1000, width = 1000) #Save file with designated name
clusterSamples(methylationInformationCov5Destrand, dist = "correlation", method = "ward", plot = TRUE) #Cluster samples based on correlation coefficients
dev.off()
```

```{r}
jpeg(filename = "2018-10-25-Loci-Analysis/2019-03-14-Full-Sample-Methylation-PCA-Cov5Destrand.jpeg", height = 1000, width = 1000) #Save file with designated name
PCASamples(methylationInformationCov5Destrand) #Run a PCA analysis on percent methylation for all samples
dev.off() #Turn off plotting device
```

```{r}
jpeg(filename = "2018-10-25-Loci-Analysis/2019-03-14-Full-Sample-Methylation-Screeplot-Cov5Destrand.jpeg", height = 1000, width = 1000) #Save file with designated name
PCASamples(methylationInformationCov5Destrand, screeplot = TRUE) #Run the PCA analysis and plot variances against PC number in a screeplot
dev.off()
```

## Obtain differentially methylated loci

```{r}
differentialMethylationStatsCov5Destrand <- calculateDiffMeth(methylationInformationCov5Destrand) #Calculate differential methylation statistics based on treatment indication from processBismarkAln
diffMethStats50Cov5Destrand <- getMethylDiff(differentialMethylationStatsCov5Destrand, difference = 50, qvalue = 0.01) #Identify loci that are at least 50% different and set a 10% FDR
head(diffMethStats50Cov5Destrand) #Confirm creation
```

```{r}
write.csv(diffMethStats50Cov5Destrand, "2018-10-25-Loci-Analysis/2019-03-13-Differentially-Methylated-Loci-Destrand-50-Cov5.csv") #Save table as .csv
```

# Create BEDfiles

## Install packages

**Do not** install these packages until analysis in `methylKit` is complete, as some of the installed packages will mask others important for analysis.

```{r}
library(readr) #Load package

install.packages("tidyverse") #Install tidyverse
library(tidyverse) #Load package
```

## 5x Destranded DMLs

```{r}
DML5xDestrand03132019 <- mutate(diffMethStats50Cov5Destrand, start = start -1, end = end + 1) %>% select(chr, start, end, meth.diff) %>% mutate_if(is.numeric, as.integer) #Save as a BED file, and avoid writing information in scientific notation
```

```{r}
head(DML5xDestrand03132019) #Confirm changes
```

```{r}
write_delim(DML5xDestrand03132019, "2019-03-13-DML-Destrand-5x-Locations.bed",  delim = '\t', col_names = FALSE) #Save data as a BED file
```
